A state space model reveals the rich and varied dynamics of replay

### 2020-07-16 Article.tex

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Department of Physiology, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}

@@

\begin{keywords}
Hippocampus | Replay | State Space
\end{keywords}

@@

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the dynamics model--how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

Finally, we specify how likely the dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the duration of the SWR, with a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

@@

Our model is specified in the context of a latent position associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

@@

\section*{Discussion}
Here we developed a state space model that identifies and quantifies generative activity in terms of a mixture of three movement dynamics: stationary, continuous, and fragmented. We showed that this model is interpretable; it has well-defined parameters and allows us to explain a large fraction of SWRs in terms of these dynamics---far greater than had been explained before. We showed that this model is flexible; it can be used with sorted or clusterless spikes, it can be used with 1D or 2D positions, and---because it uses a mixture of dynamics---it allowed us to confidently discover SWRs with non-constant speed and non-linear trajectories that would not be found with the typical Bayesian decoder. Finally, we showed that this model is robust; it was relatively insensitive to misspecified transition parameters between dynamics and depended strongly on the hippocampal place representation structure.

Using this model, we found that most SWRs contained spatially coherent content. Importantly, this content was not predominantly trajectories that occurred at many times the speed of the animal, but trajectories that were much closer to the animal's actual movement speed. Many of these events did not simply represent the animal's current location, but other locations on track. This result challenges the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999}. Are stationary events during SWR that represent the animal's position different from stationary events that represent non-local positions

Why our results are different from the Bayesian decoder
seek to describe as many SWR events, clusterless decoding, more permissive model with different dynamics, explicit descriptive model instead of implicit noisy model, smaller time bins

Our results complement those of Stella and colleagues \cite{StellaHippocampalReactivationRandom2019} who found that hippocampal replay trajectories followed Brownian-like diffusion dynamics. We found that  We also clarify how these results can coexist with Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} who found that a subset of replay trajectories alternated between stationary and jumpy representations. We also found a log normal distribution of speeds, but it should be noted that they presented their results in terms of  at least for our task, these jumpy representations are relatively rare.

Our model is similar to the competing temporal models approach of Johnson and Redish 2008 \cite{JohnsonMeasuringdistributedproperties2008}, but allows for more . Our model also represents a middle ground between Hidden Markov-style models \cite{MaboudiUncoveringtemporalstructure2018, ChenBayesiannonparametricmethods2016, LindermanBayesiannonparametricapproach2016, ChenUncoveringspatialtopology2012}, which seek to be environment-agnostic discovers of sequential patterns but require a lot of data, and typical Bayesian decoder, which has very restrictive assumptions about the nature of the trajectories. allows SWR to quickly be identified because of the discrete states. Can work in 2D. Can precisely identify timing of dynamics because of acausal decoding

We believe our model can enable a richer set of analyses that takes advantage of all the data, 
The model can easily be extend to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments.

\begin{acknowledgements}
\end{acknowledgements}

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[5]{Marielena Sosa}
\author[6]{Kenneth Kay}
\author[4]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Department of Physiology, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}
\affil[5]{Department of Neurobiology, Stanford University, Palo Alto, California}
\affil[6]{Mortimer B. Zuckerman Mind Brain Behavior Institute, Columbia University, New York, New York}

@@

\begin{keywords}
Hippocampus | Replay | State Space | Dynamics
\end{keywords}

@@

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamics model--how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the duration of the SWR, with a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

@@

\section*{Discussion}
Here we developed a state space model that identifies and quantifies generative activity in terms of a mixture of three movement dynamics: stationary, continuous, and fragmented. We showed that this model is interpretable; it has well-defined parameters and allows us to explain a large fraction of SWRs in terms of these dynamics---far greater than had been explained before. We showed that this model is flexible; it can be used with sorted or clusterless spikes, it can be used with 1D or 2D positions, and---because it uses a mixture of dynamics---it allowed us to confidently discover both SWRs with continuous many times the speed of the animal and SWRs with non-constant speed and non-linear trajectories that would not be found with the typical Bayesian decoder. Finally, we showed that this model is robust; it was relatively insensitive to misspecified transition parameters between dynamics and depended strongly on the hippocampal place representation structure.

Using this model, we found that most SWRs contained spatially coherent trajectories. Importantly, this content was not predominantly trajectories that occurred at many times the speed of the animal, but trajectories that were much closer to the animal's actual movement speed. Many of these events did not represent the animal's current location, but other locations on track. These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead there is a much richer variety of speeds the animal can use to recall prior experience or generate possible future events, including stationary representations of single locations.

Our results also explain the seeming conflict between Stella and colleagues \cite{StellaHippocampalReactivationRandom2019}, who found that hippocampal replay trajectories followed Brownian-like diffusion dynamics, and Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015}, who found that replay trajectories contained jumps in position. We found that while replay speeds are on average distributed log normally, there are still definitive "jumps" from one position to another in a small subset of the replays. Our model allows identification of both types of sequences.

We were able to find these results for several key reasons. First, we sought to describe as many SWRs as possible, instead of throwing many events out. Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated. Third, we built an explicit, but flexible model that allowed us to interpret more of the SWRs. And fourth, we used small time bins and acausal estimation of latent position, which allows us to identify the specific timing of the dynamics.

Our model can be viewed as an extension to the multiple temporal models approach of Johnson and colleagues 2008 \cite{JohnsonMeasuringdistributedproperties2008}, which used model selection to determine the model with the most appropriate trajectory speed. Our model goes beyond the Johnson model in that it explicitly permits a mixture between the movement speeds, can work for arbitrary track geometries, and uses clusterless decoding. Our model also represents a middle ground between Hidden Markov-style models \cite{MaboudiUncoveringtemporalstructure2018, ChenBayesiannonparametricmethods2016, LindermanBayesiannonparametricapproach2016, ChenUncoveringspatialtopology2012}, which seek to be environment-agnostic discovers of sequential patterns and the typical Bayesian decoder, which either uses very restrictive assumptions about the nature of the trajectories or no assumptions. The Hidden Markov-style models require lots of data and still involve interpreting the sequences in terms of known variables, such as the environment. The Bayesian decoder is either quite restrictive, when assuming a constant velocity trajectory, or potentially quite noisy, depending on the data. Both of these styles of models are dependent on the time bin size. Our model leverages some of our prior knowledge about hippocampus and its relationship to space to give less noisy and more interpretable estimates of the replay trajectories.

We believe our model can enable a richer set of analyses that takes advantage of all the data. The model can easily be extend to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments.

\begin{acknowledgements}
We thank Margaret Carr Larkin and Mattias P. Karlsson for use of their openly available datasets.
\end{acknowledgements}
